---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 6 - Due date 03/06/23"
author: "Elise Boos"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

```{r install packages}
#Load/install required package here
library(forecast)
library(tseries)
library(lubridate)
library(readxl)
library(ggplot2)
library(tseries)
library(sarima)
```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 


* AR(2)

> Answer: For an Autoregressive model the ACF will decay exponentially with time and since this is a second order AR model the pacf will identify that its second order because it will have two signifcant lag spikes on the PACF at the beginning.  

* MA(1)

> Answer: For a moving average model the PACF will decay exponentially and the ACF will identify the order, in this case as a first order moving average model if the first spike is at 1 that is order zero and there will be one additional signicant spike indicating a order of 1. 



## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

is the pacf 0.6 on the first lag?

```{r q2}

ARMAmodel_1 <- arima.sim(model=list(ar=0.6), n=100) # ar coefficient is 0.6

ARMAmodel_2 <- arima.sim(model=list(ma=0.9), n=100) # ma coefficient is 0.9

ARMAmodel_3 <- arima.sim(model=list(ar=0.6, ma=0.9), n=100)
```


(a) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r q2a}
par(mfrow=c(1,3))
Acf(ARMAmodel_1, lag = 40,main = "ACF ARMA(1,0)")
Acf(ARMAmodel_2, lag = 40,main = "ACF ARMA(0,1)")
Acf(ARMAmodel_3, lag = 40,main = "ACF ARMA(1,1)")
```


(b) Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r q2b}
par(mfrow=c(1,3))
Pacf(ARMAmodel_1, lag = 40,main = "ACF ARMA(1,0)")
Pacf(ARMAmodel_2, lag = 40,main = "ACF ARMA(0,1)")
Pacf(ARMAmodel_3, lag = 40,main = "ACF ARMA(1,1)")
```

(c) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer:

(d) Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: 


(e) Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r q2e}
ARMAmodel_1 <- arima.sim(model=list(ar=0.6), n=1000) # ar coefficient is 0.6

ARMAmodel_2 <- arima.sim(model=list(ma=0.9), n=1000) # ma coefficient is 0.9

ARMAmodel_3 <- arima.sim(model=list(ar=0.6, ma=0.9), n=1000)

par(mfrow=c(1,3))
Acf(ARMAmodel_1, lag = 40,main = "ACF ARMA(1,0)")
Acf(ARMAmodel_2, lag = 40,main = "ACF ARMA(0,1)")
Acf(ARMAmodel_3, lag = 40,main = "ACF ARMA(1,1)")

par(mfrow=c(1,3))
Pacf(ARMAmodel_1, lag = 40,main = "ACF ARMA(1,0)")
Pacf(ARMAmodel_2, lag = 40,main = "ACF ARMA(0,1)")
Pacf(ARMAmodel_3, lag = 40,main = "ACF ARMA(1,1)")
```



## Q3

assumption 0 or 1 depending on having or not having the constant in the equation 

if no constant term the series has been differenced 

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

identify the integers in the model 
(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

(b) Also from the equation what are the values of the parameters, i.e., model coefficients. 


## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r q4}
par(mfrow=c(1,2))
SARIMAmodel_4 <- sim_sarima(model=list(ma=0.5,sar=0.8,nseasons=12), n=1000)
Acf(SARIMAmodel_4, lag =40)
Pacf(SARIMAmodel_4, lag =40)
```

