---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Elise Boos"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
require(forecast)
require(tseries)
require(Kendall)
require(lubridate)
require(tidyverse)

#load data
energydat <- read.csv("./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv", 
                      stringsAsFactors = TRUE)

#select columns
energydat <- energydat %>% 
  select(Month, Total.Biomass.Energy.Production, 
         Total.Renewable.Energy.Production, Hydroelectric.Power.Consumption)
#set to date object 
energydat$Month <- ym(energydat$Month)
head(energydat_sel)
```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r question 1}
#run time series 
ts_energy <- ts(energydat[2:4], 
                start = c(year(first(energydat$Month)),
                                              month(first(energydat$Month))), 
                frequency = 12)

par(mfrow = c(1, 3))
for (i in 1:3){
  plot(ts_energy[,i]) +
  title(main=paste(colnames(ts_energy)[i]))
  acf(ts_energy[,i], lag.max =40) 
  pacf(ts_energy[,i], lag.max =40)
}

```

### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Total biomass energy production and Total Renewable Energy production appear to have an upward trend while Hydroelectric Power Consumption appears to have a downward trend.


### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r question 3}
nobs <- nrow(energydat)
t <- c(1:nobs) 

#trend on biomass
linear_trend_bio <- lm(energydat[,2]~t)  
summary(linear_trend_bio)
#save coefficients
beta0_bio=as.numeric(linear_trend_bio$coefficients[1])  #intercept 
beta1_bio=as.numeric(linear_trend_bio$coefficients[2])  #slope

#trend on renewable
linear_trend_ren <- lm(energydat[,3]~t)  
summary(linear_trend_ren)
#save coefficients
beta0_ren=as.numeric(linear_trend_ren$coefficients[1])  #intercept 
beta1_ren=as.numeric(linear_trend_ren$coefficients[2])  #slope

#trend on hydro
linear_trend_hydro<- lm(energydat[,4]~t)  
summary(linear_trend_hydro)
#save coefficients
beta0_hydro=as.numeric(linear_trend_hydro$coefficients[1])  #intercept 
beta1_hydro=as.numeric(linear_trend_hydro$coefficients[2])  #slope

```
 
####INTERPRETAITON  
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r question 4}
detrend_bio <- energydat[,2]-(beta0_bio+beta1_bio*t)
detrend_ren <- energydat[,3]-(beta0_ren+beta1_ren*t)
detrend_hydro <- energydat[,4]-(beta0_hydro+beta1_hydro*t)

ggplot(energydat, aes(x=Month, y=energydat[,2])) +
            geom_line(color="blue") +
            ylab(paste(colnames(energydat)[2])) +
            geom_smooth(color="black",method="lm") +
            geom_line(aes(y=detrend_bio), col="lightblue")+
            geom_smooth(aes(y=detrend_bio),color="black",method="lm") 

ggplot(energydat, aes(x=Month, y=energydat[,3])) +
            geom_line(color="blue") +
            ylab(paste(colnames(energydat)[3])) +
            geom_smooth(color="black",method="lm") +
            geom_line(aes(y=detrend_ren), col="lightblue")+
            geom_smooth(aes(y=detrend_ren),color="black",method="lm") 

ggplot(energydat, aes(x=Month, y=energydat[,4])) +
            geom_line(color="blue") +
            ylab(paste(colnames(energydat)[4])) +
            geom_smooth(color="black",method="lm") +
            geom_line(aes(y=detrend_hydro), col="lightblue")+
            geom_smooth(aes(y=detrend_hydro),color="black",method="lm") 

```

###INTERPRETATION

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

```{r question 5}
par(mfrow = c(1, 2))
acf(detrend_bio, lag.max =40) 
pacf(detrend_bio, lag.max =40)

acf(detrend_ren, lag.max =40) 
pacf(detrend_ren, lag.max =40)

acf(detrend_hydro, lag.max =40) 
pacf(detrend_hydro, lag.max =40)
```

####INTERPRETATION

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

```{r question 6}
dummies <- seasonaldummy(ts_energy[,1])

#Then fit a linear model to the seasonal dummies
seas_model_bio=lm(energydat[,2]~dummies)
summary(seas_model_bio)

seas_model_ren=lm(energydat[,3]~dummies)
summary(seas_model_ren)

seas_model_hydro=lm(energydat[,4]~dummies)
summary(seas_model_hydro)

#Look at the regression coefficient. These will be the values of Beta

#Store regression coefficients
beta_int_bio=seas_model_bio$coefficients[1]
beta_coeff_bio=seas_model_bio$coefficients[2:12]

beta_int_ren=seas_model_ren$coefficients[1]
beta_coeff_ren=seas_model_ren$coefficients[2:12]

beta_int_hydro=seas_model_hydro$coefficients[1]
beta_coeff_hydro=seas_model_hydro$coefficients[2:12]
```

####INTERPRETATION

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r question 7}
energy_seas_comp_bio=array(0,nrow(energydat))
for(i in 1:nobs){
  energy_seas_comp_bio[i]=(beta_int_bio+beta_coeff_bio%*%dummies[i,])
}

deseason_energy_bio <- energydat[,2]-energy_seas_comp_bio

ggplot(energydat, aes(x=Month, y=energydat[,2])) +
            geom_line(color="blue") +
            ylab(colnames(energydat)[(2)]) +
            geom_line(aes(y=deseason_energy_bio), col="green")

energy_seas_comp_ren=array(0,nrow(energydat))
for(i in 1:nobs){
  energy_seas_comp_ren[i]=(beta_int_ren+beta_coeff_ren%*%dummies[i,])
}

deseason_energy_ren <- energydat[,3]-energy_seas_comp_ren

ggplot(energydat, aes(x=Month, y=energydat[,3])) +
            geom_line(color="blue") +
            ylab(colnames(energydat)[(3)]) +
            geom_line(aes(y=deseason_energy_ren), col="green")

energy_seas_comp_hydro=array(0,nrow(energydat))
for(i in 1:nobs){
  energy_seas_comp_hydro[i]=(beta_int_hydro+beta_coeff_hydro%*%dummies[i,])
}

deseason_energy_hydro <- energydat[,4]-energy_seas_comp_hydro

ggplot(energydat, aes(x=Month, y=energydat[,4])) +
            geom_line(color="blue") +
            ylab(colnames(energydat)[(4)]) +
            geom_line(aes(y=deseason_energy_hydro), col="green")
```

###INTERPRETATION

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?

```{r question 8}
par(mfrow = c(1, 2))
acf(deseason_energy_bio, lag.max =40) 
pacf(deseason_energy_bio, lag.max =40)

acf(deseason_energy_ren, lag.max =40) 
pacf(deseason_energy_ren, lag.max =40)

acf(deseason_energy_hydro, lag.max =40) 
pacf(deseason_energy_hydro, lag.max =40)

```


